---
title: "ML_Proj8 - Final"
author: "Austin Marckx"
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(GGally)
library(ggpubr)
library(readxl)
library(randomForest)
library(caret)
library(cowplot)
library(tidyverse)
library(ggridges)
library(boot)
library(mclust)
library(corrplot)
library("FactoMineR")
library("factoextra")
library(nlme)
library(lme4)
library("REEMtree")
library(brms)
library(rstatix)

minmax <- function(x){
  return( (x - min(x))/(max(x) - min(x))  )
}

set.seed(3141526)
```

```{r}
# Read in file
df <- read_xlsx('./data/AD.training.xlsx')

# Transform datatypes
df <- df %>% transform(
  PTID = factor(PTID),
  DX = factor(DX.bl, levels = c("CN", "EMCI", "LMCI", "AD")),
  PTGENDER = factor(PTGENDER),
  APOE4 = factor(APOE4)
) %>% select(AGE, PTEDUCAT, MMSE, PTGENDER, APOE4, DX, everything(), -DX.bl)

# Summary
df %>% head()
df %>% str()
df %>% summary()

any(is.na(df))
df[!complete.cases(df),]

# Train test split:
df <- df %>% select(-RID, -EXAMDATE)
train <- df %>% dplyr::sample_frac(0.70)
test  <- dplyr::anti_join(df, train, by = 'PTID')

```

# GLM Regression

## gaussian
```{r}
gauss <- glm(`MMSE.Change` ~ AGE+MMSE+PTGENDER+PTEDUCAT+APOE4+DX, data = train, family = gaussian())
coef(summary(gauss))[,4]
summary(gauss)

# RMSE
sqrt(mean(gauss$residuals ^ 2))
plot(gauss)

train$glm_train <- predict(gauss)
test$glm_test <- predict(gauss, newdata = test)

```


# Random Forest Regression: 
```{r}
rf <- randomForest(`MMSE.Change` ~ AGE+MMSE+PTGENDER+PTEDUCAT+APOE4+DX, data = train, ntrees = 500, importance = TRUE, type = 'regression')

# RMSE
sqrt(mean(rf$mse))

rf
plot(rf)
importance(rf)
varImpPlot(rf)

train$rf_train <- predict(rf)
test$rf_test <- predict(rf, newdata = test)

```

# Linear Mixed Effects

```{r}
linMixEff <- lme(`MMSE.Change`~ AGE+MMSE+PTGENDER+PTEDUCAT+APOE4+DX, random = ~ 1|MMSE/DX, data = train)
linMixEff
print("RMSE: ")
sqrt(mean(linMixEff$residuals ^ 2))
summary(linMixEff)

plot(linMixEff)
plot(ranef(linMixEff))

train$linMixEff_train <- predict(linMixEff)
test$linMixEff_test <- predict(linMixEff, newdata = test)

```

# Bayesian Mixed Effects Regression: 
```{r}
bme <- brm(MMSE.Change ~ AGE+MMSE+PTGENDER+PTEDUCAT+APOE4+DX + (1|DX), data = train)
summary(bme)
pp_check(bme, ndraws = 30)

# Get preds
train$bme_train <- predict(bme)[,1]
test$bme_test <- predict(bme, newdata = test)[,1]

```
# RMSE:
```{r}
train <- train %>%
  pivot_longer(cols = c(glm_train, rf_train, linMixEff_train, bme_train), names_to = "model", values_to = "ypred") %>%
  mutate(predresidual = MMSE.Change - ypred, model = factor(model, levels = c("rf_train","glm_train", "linMixEff_train", "bme_train")) ) #%>%

test <- test %>%
  pivot_longer(cols = c(glm_test, rf_test, linMixEff_test, bme_test), names_to = "model", values_to = "ypred") %>%
  mutate(predresidual = MMSE.Change - ypred, model = factor(model, levels = c("rf_test","glm_test", "linMixEff_test", "bme_test")) )

results <- train %>% bind_rows(test)
```



```{r}
# RMSE:
results %>% 
  group_by(model) %>%
  summarize(RMSE = sqrt(mean(ypred ^ 2)))

```



# Plot the evals/predictions
```{r}
results %>% 
  ggplot(aes(x = model, y = sqrt(abs(predresidual)), color = model, fill = model)) + 
    geom_jitter(width = 0.3, height = 0.035, alpha = 0.7) +
    stat_summary(fun.y = mean, geom = "bar", alpha = 0.15) +
    stat_summary(fun.data = mean_se, geom = "errorbar", color = "black", width= 0.2) +
    theme_minimal()

```

```{r, fig.height = 12, fig.width = 12}
train %>% 
  ggplot(aes(x = model, y = sqrt(abs(predresidual)), color = model, fill = model)) + 
    geom_point(alpha = 0.5) +
    geom_line(aes(group = PTID), size = 0.01) +
    scale_y_log10() + 
    theme_minimal() 

test %>% 
  ggplot(aes(x = model, y = sqrt(abs(predresidual)), color = model, fill = model)) + 
    geom_point(alpha = 0.5) +
    geom_line(aes(group = PTID), size = 0.01) +
    scale_y_log10() + 
    theme_minimal() 
```


```{r}
train %>% 
  ggplot(aes(x = MMSE.Change, y = ypred, color = model)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, color = "red") +
  theme_minimal() + 
  facet_wrap(~model)


test %>%
  ggplot(aes(x = MMSE.Change, y = ypred, color = model)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, color = "red") +
  theme_minimal() + 
  facet_wrap(~model)

```


```{r}
sessionInfo()
```

Archive:
train %>% 
  ggplot(aes(x = model, y = sqrt(abs(predresidual)), color = model)) + 
    geom_jitter(width = 0.3, height = 0.035, alpha = 0.7) +
    stat_summary(fun.data = mean_se, geom = "errorbar", color = "black", width= 0.2) +
    theme_minimal()

test %>% 
  ggplot(aes(x = model, y = sqrt(abs(predresidual)), color = model)) + 
    geom_jitter(width = 0.3, height = 0.035, alpha = 0.7) +
    stat_summary(fun.data = mean_se, geom = "errorbar", color = "black", width= 0.2) +
    theme_minimal()

